---
title: 'Day 1 activity: UN Votes'
author: "YOUR NAME HERE"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

*This activity is adapted from [Data Science in a Box](https://datasciencebox.org) by Mine Çetinkaya-Rundel*

## Introduction

How do various countries vote in the United Nations General Assembly, how have their voting patterns evolved throughout time, and how similarly or differently do they view certain issues?
Answering these questions (at a high level) is the focus of this analysis.

This document is an "Rmd" or "R markdown" file.

```{r}
2+2
x <- 3
x^2
```


### Packages

R is open source and user contributed!
Different functionalities exist in *packages* which you access each time you start a new document or R session.
We will use the **tidyverse** and **lubridate** packages for data wrangling and visualization, and the **DT** package for interactive display of tabular output, and the **unvotes** package for the data.
We use **ggthemes** to modify the visual appearance of the graph, including color palettes.

```{r}
#| label: load-packages
#| warning: false
#| message: false

library(tidyverse)
library(lubridate)
library(DT)
library(unvotes)
library(ggthemes)

```

### Data

The data we're using originally come from the **unvotes** package.
In the chunk below we modify the data by joining the various data frames provided in the package to help you get started with the analysis.
*(We'll see this in Chapter 20, around Day 15)*.

```{r}
unvotes <- un_votes %>%
  inner_join(un_roll_calls, by = "rcid") %>%
  inner_join(un_roll_call_issues, by = "rcid")

unvotes
```

## UN voting patterns

Let's create a data visualization that displays how the voting record of the United States changed over time on a variety of issues and compare it to two other countries: the UK and Turkey.

We can easily change which countries are being plotted by changing which countries the code above `filter`s for.
Note that the country name should be spelled and capitalized exactly the same way as it appears in the data.
See the [Appendix](#appendix) for a list of the countries in the data.

*We will learn `filter` on Day 5 (Ch 3.2), `mutate` on Day 6 (Ch 3.3) and `group_by` and `summarize` on Day 10 (Ch 3.5).*

*We will start by learning more about creating plots with `ggplot` on Days 1-4!*

```{r}
#| label: plot-yearly-yes-issue
#| message: false
#| fig-width: 10
#| fig-height: 6
unvotes %>% 
  filter(country %in% c("United Kingdom", 
                        "United States", 
                        "Turkey")) %>%
  mutate(year = year(date)) %>%
  group_by(country, year, issue) %>%
  summarize(percent_yes = mean(vote == "yes")) %>%
  ggplot(mapping = aes(x = year, 
                       y = percent_yes, 
                       color = country)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~issue) +
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  labs(
    title = "Percentage of 'Yes' votes in the UN General Assembly",
    subtitle = "1946 to 2019",
    y = "% Yes",
    x = "Year",
    color = "Country"
  ) +
  theme_bw() 
```

# Working together we will:

-   Run each chunk! Use the run button or the keyboard shortcut "Cmd/Ctrl + Enter/Return"
-   Add `scale_color_colorblind()` to make this chart more accessible!
-   Change the last line, `theme_bw()`, to `theme_clean()`. What happens? There are lots of themes to explore! Hit tab after you type `theme_` to see more options.
-   Try commenting out the line `scale_y_continuous(...)` to see what it does. (Add \# at the beginning of the line and then re run the chunk)
-   Comment out multiple lines to see what "labs" does. Highlight lines 76 - 82, then hit "Cmd/Ctrl + shift + C". Then re-run the chunk!
-   Save the Rmd file to your "In Class Work" subfolder in your Student folder.
-   Knit the document!

# Your turn!

1.  Pick three countries of your own choosing!
    (Run the code in Appendix below for a list).
    Describe patterns that you see.
    Which countries vote most similarly?
    On which issues?
    Did this change over time?

2.  Try a new theme to alter the general visual display of the graph!
    Use the tab key to see all your options!

3.  Before you leave class, be sure the document will still Knit!

## References

1.  David Robinson (2017). [unvotes](https://CRAN.R-project.org/package=unvotes): United Nations General Assembly Voting Data. R package version 0.2.0.
2.  Erik Voeten "Data and Analyses of Voting in the UN General Assembly" Routledge Handbook of International Organization, edited by Bob Reinalda (published May 27, 2013).
3.  Much of the analysis has been modeled on the examples presented in the [unvotes package vignette](https://cran.r-project.org/web/packages/unvotes/vignettes/unvotes.html).

## Appendix {#appendix}

Below is a list of countries in the dataset:

```{r, eval=FALSE}
#| label: list-countries
#| echo: false

unvotes %>% 
  select(country) %>%
  arrange(country) %>% 
  distinct() %>%
  datatable()
```

## A few terms you heard today:

*R* is the programming language, the code.

*RStudio* is the interface - all the windows and buttons we use to make things go.

Sometimes you'll hear me (or other people) use the words interchangeably, to mean "use R to do stuff in RStudio!" :)

The gray boxes are called code *chunks*, and everything inside them is *R* code.

This file is an *R Markdown* document.
The symbols you use to change bold, italics, and other formatting are *Markdown* syntax.
The end of the file is *.Rmd*, that's how RStudio knows to treat it as a *R Markdown* file.
We will always use this type of file in the class.
Everything outside the code chunks (not in gray) is text (or formatting symbols).
