---
title: "Transformation on rows"
subtitle: "filter, arrange, distinct, count"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
editor_options:
  chunk_output_type: console
---

## What are Tidy Data?

Three rules which make a data set "tidy":

1. each variable must have its own column
2. each observation must have its own row
3. each value must have its own cell

Why use tidy data?

- consistent data structure important
- variables in columns allows R's vectorized nature to shine

```{r}
# Different ways of showing TB cases documented by WHO
table1  # tidy!
table2  # breaks rule 1
table3  # breaks rule 3
table4a # breaks rule 2
table4b # breaks rule 2 
```



## This code should be used with Class > Code > Workshop_Day1 > Images > filter_inclass_activity.pdf.

```{r}
library(tidyverse)
```


```{r}
data <- tibble(
  SDScon = c("Yes", "No", "Yes", "No", NA), 
  pets = c(0,2,1,0,1), 
  pics = c(8742, 4536, 10119, 403, 31492), 
  fydorm = c("H", "K", "E", "M", "K")
)
```


```{r}
data
```


```{r}
data |>
  filter(pics < 1000)


data |>
  filter(pets >= 1)


data |>
  filter(SDScon == "Yes")


data |>
  filter(SDScon != "Yes")


data |> 
  filter(fydorm %in% c("K", "E"))


data |>
  filter(is.na(SDScon))


data |> 
  filter(SDScon == "Yes" | is.na(SDScon))



data |> 
  arrange(pets)


data |>
  arrange(fydorm)


data |>
  distinct(SDScon)


data |>
  distinct(pets)


data |>
  count(SDScon)


data |>
  count(pets)


# count() gives the number of obs in each category, alphabetical order with NAs on the bottom  

# distinct() gives all the unique values in a column, in the order they were in original df 

# %in% c() #match rows in this list where c() makes a list 

# & is AND and | is OR

```


**So, why do we need `is.na(x)` rather than just `filter(x != NA)`?**

Remember that NA values are *contagious*... any time an NA is an input, NA will be the output. 

```{r}
1 + NA
3*NA

testdat <- tibble(x = c(1:2, NA, NA, 3), y = 1:5)
testdat

testdat |> 
  mutate(
    x2 = 2*x, 
    x_plus_y = x + y, 
    x_gt2 = x>2
  )
```

Watch what happens with NA:  (HINT: run one line at a time with Ctrl + Enter)

```{r}
1 == 1
1 == NA
NA == NA

1 != NA
NA != NA


testdat |>
  mutate(
    equalNA = (x == NA),
    notequalNA = (x != NA), 
    isna = is.na(x), 
    not_isna = !is.na(x)
  )

testdat |> 
  filter(x == NA)

testdat |>
  filter(is.na(x))

testdat |> 
  filter(x != NA)

testdat |>
  filter(!is.na(x))
```

