---
title: "transformations on columns KEY"
author: "select, rename, mutate, relocate"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---


## Learning Goals:

- use the mutate, select, rename, and relocate functions to complete one-step data tasks

- use <- as the assignment operator

- identify syntactic and non-syntactic variable names

- use ` ` backticks to surround non-syntactic variable names

------------------------------------------------------------------------

```{r}
library(tidyverse)

bball_full <- read_csv("~/Sds_101A_SSI26/Class/Data/basketball_2023_2.csv")

bball_full
```


# select()

Use select to pick one column:

```{r}
bball_full |>
  select(Player)
```

More than one column:

```{r}
bball_full |>
  select(Player, Pos, Tm)
```

Or a bunch of columns:

```{r}
bball_full

bball_full |>
  select(Rk:FG)
```


1. Select the columns Player, Tm, G, MP, and FG through `3P%`


### Non-syntactic names

Wait a minute... why do we have `3P%` instead of just 3P%? There are rules about what R will accept as variable names. 

**syntactic** names (names that "follow the rules")

- start with a letter (a-z or A-Z), NOT a number or punctuation.

- contain only letters, numbers, _ or .

- do not contain spaces or special characters (other than _ or .)


**non-syntactic** names break at least one of the above rules.

When referring to non-syntactic variable names, you must surround them by "back ticks" (I also call these special quotation marks). 


2. Would "Player Name" be a syntactic or non-syntactic name? Why?


### Here are some examples we examined in class

In ggplot (and most other places!) you MUST use backticks around non-syntactic names.

```{r}
ggplot(bball_full) + 
  geom_point(aes(x= MP, y = `3PA`))
```


```{r}
ggplot(bball_full) + 
  geom_point(aes(x= MP, y = "3PA"))
```

BUT....

select and rename allow you to refer to ALL variables (syntactic or not) with quotations. So you might have noticed that regular double " or single ' quotes also worked in select above.

```{r}
bball_full |>
  select(Player, Tm, G, MP, FG:"3P%")
```



### <- The assignment operator

OK, so lets look at this:

```{r}
bball_full |> 
  select(Player, Age)
```

Notice that all the code we ran before just PRINTED OUT the results of the select. It made no permanent changes to the dataset bball_full.

The example below shows how the assignment operator works. You also saw this in homework and in past classes.

```{r}
bball_ages <- bball_full |>  #creates a new dataframe
  select(Player, Age)
```

Above, we created a new dataframe in our environment called bball_ages using our assignment operator (<-). How many rows does it have? How many columns? Compare those numbers to our original bball_full dataset. 

```
NEW_DATASET <- OLD_DATASET |> select(blah, blah, blah)
```

3.  Select the columns Player, Tm, G, MP, and FG through `3P%`, and save this smaller dataset as "bball_small".  

Type the name bball_small to print it out!


[after your attempt scroll down for the answer]














```{r}
# This part creates bball_small: 
bball_small <- bball_full |>
  select(Player, Tm, G, MP, FG:`3P%`)


# This part prints it out:
bball_small
```


# rename()

As you might suspect, this function renames columns! Remember: new name = old name.

```{r}
bball_small |>
  rename(Team = Tm)
```


```{r}
bball_small |>
  rename(field_goal_pct = `FG%`)
```

4. Look at the output from the rename above. Why is the Team column still say Tm, not Team? 


We can also rename more than one column at a time!

```{r}
bball_small <- bball_small |>
  rename(
    Team = Tm,
    field_goal_pct = `FG%`
  )
```


# mutate()

The mutate function is used to CHANGE or ADD new columns to your dataset that are based on existing columns.  What does the code below do? Write it out line by line. 

```{r}
bball_small |>
  mutate(three_point_pct = `3P%` * 100) |>
  select(Player, Team, `3P%`, three_point_pct)
```

If you reuse an existing name, it will replace that column:

```{r}
bball_small |>
  mutate(`3P%` = `3P%` * 100) |>
  select(Player, Team, `3P%`)
```


Your mutate can also depend on multiple variables:

```{r}
bball_small |>
  mutate(FG_att_per_minute = FGA / MP) |>
  select(Player, Team, FGA, MP, FG_att_per_minute)
```


For each of the following, start with **bball_small**!

5. Basketball stats are often reported per each 36 minutes played. Create a column, fga_per_36, which is the field goal attempts per 36 minutes. 


6. Create a column to verify that the `3P%` variable is the number of 3P (3 points made) divided by 3PA (3 point attempts).


### mutating multiple variables at once

We can also mutate more than one variable at a time. Separate each with a comma. 

You can also refer to new variables you have created! Each column is created in order.

```{r}
bball_small |>
  rename(three_pt_attempts = `3PA`) |>
  mutate(
    three_att_per_min = three_pt_attempts / MP,
    three_att_per_36 = three_att_per_min * 36,
    fg_per_min = FG / MP
  ) |>
  select(Player, Team, three_pt_attempts, MP, three_att_per_min,
         three_att_per_36, FG, fg_per_min)
```


# relocate()

This rearranges the order of the columns. By default, it puts the listed variable(s) first and then preserves the order of the remaining columns.

```{r}
bball_small |> 
  relocate(Team, MP)
```

There are also options, .before and .after you can use:

```{r}
bball_small |>
  relocate(MP, .before = Team)

bball_small |>
  relocate(MP, .after = Team)
```


## Extra Practice! 

If your table has finished the above, try this extra practice problem that combines our knowledge about transforming rows AND transforming columns! 

Problem: 

Start with `bball_full` and create a new dataset called `bball_extra`. In this data set, I want only point guards (PG) who are older than 22 years old. I would like a new column that adds the number of two-point and three-point shots made called `shots_total`. I am only interested in the columns `Pos`, `Age`, and `shots_total`. Finally I would like a scatterplot of `Age` as the explanatory variable and  `shots_total` as the response, with a smoother added. Can you do this all in one code chunk?  What can you conclude from your plot?

