---
title: "Exploratory Data Analysis"
output:
  pdf_document:
    fig_height: 3
    fig_width: 4.5
  word_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Summarizing and visualizing recap

The following table recaps the methods for summarizing and visualizing two variables

| Explanatory | Response    | Summarize           | Visualize                   |
|-------------|:------------|:--------------------|:----------------------------|
| None        | Categorical | table, prop table   | Bar plot                    |
| Categorical | Categorical | table, prop table   | Bar plot, Mosaic plot       |
| None        | Numerical   | summary stats       | Histogram, Box plot         |
| Categorical | Numerical   | summary stats table | Stacked Histogram, Box plot |
| Numerical   | Numerical   | correlation         | Scatter plot                |

In a similar way, the following table summarizes the key `R` commands:

| Explanatory | Response    | Summarize                                  | Visualize                                        |
|-------------|:------------|:-------------------------------------------|:-------------------------------------------------|
| None        | Categorical | `table()`, `proportions()`                 | `geom_bar()`                                     |
| Categorical | Categorical | `table()`, `proportions()`, `addmargins()` | `geom_bar(position = "fill")` `geom_mosaic()`    |
| None        | Numerical   | `mosaic:favstats()`                        | `geom_boxplot()` `geom_histogram()`              |
| Categorical | Numerical   | `mosaic:favstats()`                        | `geom_boxplot()` `geom_histogram()+facet_grid()` |
| Numerical   | Numerical   | `cor()`                                    | `geom_point() + geom_smooth()`                   |

-   Notice commands for summarizing are put together using `%>%` (pipe)

-   Notice commands for plots are put together using `+`

-   The notation `y~x` means `y` is the response variable and `x` is the explanatory variable. We can use this in `table()` and in `favstats()`

\newpage

# Loading and Querying our dataset

-   Loading the dataset:

```{r}
library(tidyverse)
boybands <- read_csv("~/Sds_172_S26/Class/Data/boybands.csv")
pockets <- read_csv("~/Sds_172_S26/Class/Data/pockets.csv")
```

-   Calculating the size of our dataset:

```{r}
dim(boybands)
```

-   Getting the names of the variables in our dataset

```{r}
colnames(pockets)
```

\newpage

# Single categorical variable

## Summarize

-   Table:

```{r}
table(boybands$hair_length)
```

-   Proportions table with 2 decimal number

```{r}
table(boybands$hair_length) %>% proportions() %>% round(2)
```

## Visualize

-   Bar plot:

```{r}
ggplot(boybands, aes(x=hair_length))+
  geom_bar()+
  labs(x="Hair Length",
       title = "Most boyband members have short hair")
```

\newpage

# Two categorical variables

## Summarize:

-   Table:

```{r}
table(boybands$hair_length, boybands$hair_frosted)
```

-   Table with totals per row and per column

```{r}
table(boybands$hair_length, boybands$hair_frosted) %>% addmargins()
```

-   Proportions table, with proportions per row

```{r}
table(boybands$hair_length, boybands$hair_frosted) %>% proportions(margin=1) %>% round(3)
```

## Visualize:

-   Stacked bar plot

```{r}
ggplot(data = boybands, aes(x = hair_length, fill = hair_frosted)) + 
  geom_bar() +
  labs(x = "Hair Length",
       title = "No frosting on long hair",
       fill = "Frosted tips?")
```

-   Proportions bar plot:

```{r}
ggplot(data = boybands, aes(x = hair_length, fill = hair_frosted)) + 
  geom_bar(position="fill") +
  labs(x = "Hair Length",
       title = "No frosting on long hair",
       fill = "Frosted tips?")
```

-   Mosaic plot:

```{r,eval=FALSE}
library(ggmosaic)
ggplot(data = boybands) + 
  geom_mosaic(aes(x = product(hair_length), fill = hair_frosted))+
  labs(x = "Hair Length",
       title = "No frosting on long hair",
       fill = "Frosted tips?")

```

\newpage

# Single numerical variable

## Summarize

-   Summary statistics

```{r}
mosaic::favstats(pockets$maxHeightFront)
```

Remember to describe using the mnemonic `CUSS`, which stands for:

-   *C*enter: Mean or median.
-   *U*nusual features: Outliers.
-   *S*pread: Standard Deviation, Interquartile range.
-   *S*hape: Symmetric, skewed, multimodal

## Visualize

-   Histogram

```{r}
ggplot(pockets, aes(x = maxHeightFront)) +
  geom_histogram(bins = 10) +
  labs(x = "Maximum height of front pockets (cm)",
       title = "Jeans' pocket heights")
```

-   Boxplot

```{r}
ggplot(pockets, aes(x = maxHeightFront)) +
  geom_boxplot() +
  labs(x = "Maximum height of front pockets (cm)",
       title = "Jeans' pocket heights")
```

\newpage

# Categorical by numerical

## Summarize

-   Summary statistics:

```{r}
mosaic::favstats(maxHeightFront~menWomen, data = pockets)
```

## Visualize:

-   Boxplot:

```{r}
ggplot(pockets, aes(x=maxHeightFront, y=menWomen))+
  geom_boxplot()+
  labs(x="Height of Front Pocket",
     title="Women's pockets are inferior")
```

-   Stacked histograms:

```{r}
ggplot(pockets, aes(x=maxHeightFront))+
  geom_histogram(bins=10)+
  facet_grid(menWomen~.)+
  labs(x="Height of Front Pocket",
       title="Women's pockets are inferior")
```

\newpage

# Two numerical variables

## Summarize

-   Calculate correlation

```{r}
mosaic::cor(maxHeightFront ~ price, data = pockets)
```

## Visualize:

-   Scatterplot

```{r}
ggplot(pockets, aes(x = price, y = maxHeightFront)) + 
  geom_point() + 
  labs(x = "Price", y = "Height of Front Pocket")
```

-   Scatterplot with linear trend

```{r}
ggplot(pockets, aes(x = price, y = maxHeightFront)) + 
  geom_point() + 
  geom_smooth(method = lm) +
  labs(x = "Price", y = "Height of Front Pocket")

```
