---
title: 'Review of Statistics 1 Material'
subtitle: 'A quick review of linear modeling'
author: "M Zillig"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

## Learning Objectives

Here, we will review functions and statistical methods you learned in Statistics 1. Specifically, we will review how to summarize and visualize numeric data and how to run as simple linear model.

## New Functions

-   `lm()` Runs a linear regression

-   `ggplot()` Creates a plot

## Set Up

```{r, message = F, warning = F}
library(tidyverse)

porsche <- read_csv("~/Sds_101A_SSI26/Class/Code/Workshop_Day3/PorschePrice.csv")
```

## Before Class:

Review Chapters 3-22 in your textbook from Statistics 1.

## During Class

### Example 1.1: Porsche price vs. mileage

First, let's examine our data. We are interested in how the price of a porsche changes based on the mileage.

```{r}
#Create a summary of the three variables 
summary(porsche)


#create a scatterplot showing the relationship between mileage and price 

ggplot(data = porsche, aes(x = Mileage, y = Price))+
  geom_point()

```

Next, let's run a simple linear model:

```{r}
porsche_fit <- lm(data = porsche, Price ~ Mileage)

summary(porsche_fit)

```

1.  What is the slope estimate? Give an interpretation of the slope in the context of the question.

2.  What is the R-squared value? Give an interpretation of the R-squared value in the context of the question.

Because we only have a single slope and intercept term, we can easily use `geom_smooth()` to add the line of best fit:

```{r}

ggplot(data = porsche, aes(x = Mileage, y = Price))+
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

```
