---
title: "Summarizing data"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

# Learning Goals:

-   Use `summarize()` to create tables of summary statistics.

-   Group by one or more variables before `summarize()`


------------------------------------------------------------------------

```{r, message = FALSE}
library(tidyverse)
library(palmerpenguins)
```

# Summarizing data


## Loading our data


Let's start by loading the results of a student survey, that contains information about their favorite food, their major, and the number of books they read during last year

```{r}
(survey <- read_csv("~/Sds_101A_SSI26/Class/Data/student_survey.csv"))
```

## Some quick summaries

Initially we are interested in learning the average amount of books in our survey:

```{r}
survey |>
  summarize(avg_books=mean(num_books))
```

The key function in this case is `summarize()` which creates a new column (`avg_books`), by applying the function `mean()` on all the values of the column `num_books`. Notice in general the left side corresponds to the column name, while the right side corresponds to the type of summary we are using.

Let's alter our previous code to create a couple of different summaries:

* The number of observations in our survey

```{r}
survey |>
  summarize(num=n())

```


* The minimum number of books read in our survey:

```{r}
survey |>
  summarize(min_books=min(num_books))
```


* The maximum number of books read in our survey:

```{r}
survey |>
  summarize(max_books=max(num_books))
```

* The median number of books read in our survey:

```{r}
survey |>
  summarize(med_books=median(num_books))
```


And notice that we can combine all the previous summaries in the following statement

```{r}
survey |>
  summarize(
    num = n(),
    avg_books=mean(num_books),
    min_books=min(num_books),
    max_books=max(num_books),
    med_books=median(num_books)
    )
```

## Using groups

Notice that the previous analysis was applied to all the observations in our dataset. Now let's write code to find the average number of books read across each of our majors

```{r}
survey |>
  group_by(major)|>
  summarize (avg_books = mean(num_books))
```

Notice that we added the command `group_by(major)` which group all the observations with a similar major and then applies the average across them.

# Your turn 1: Predicting rows and columns

In your groups, predict the number of rows and columns that the following code will produce. 

**Make sure to record your prediction before you run the code**. 

**After you record your predictions compare them by running the code**

1. **YOUR GUESS**, Number of rows = xxx, Number of columns == yyy

```{r}
survey |>
  group_by(fav_food) |> 
  summarize(
    num_students = n(),
    max_num_books = max(num_books)
  )
```


2. **YOUR GUESS**, Number of rows = xxx, Number of columns == yyy

```{r}
survey |> 
  group_by(major) |> 
    summarize(
      num_students = n(),
      max_num_books = max(num_books)
    )

```

3. **YOUR GUESS**, Number of rows = xxx, Number of columns == yyy

```{r}
survey |> 
  group_by(major, fav_food) |> 
    summarize(num_students = n())
```


# Using the Palmer Penguins! 

Let's use these data again! :)

```{r}
penguins <- palmerpenguins::penguins
```

To practice what we learnt, let's calculate how many penguins from each species we have

```{r}
penguins |>
  group_by(species) |>
  summarize(n=n())
```

And now let's calculate the average body mass (in grams) across each of the species:

```{r}
penguins |>
  group_by(species) |>
  summarize(
    n=n(),
    avg_weight=mean(body_mass_g))
```


# Your turn 2: Dealing with missing observations

1. In your groups discuss what was surprising about the previous output. Any ideas of why you run into issues?

2. Let's display our dataset using:

```{r}
penguins
```

Are there observations (row) where there is missing values in `body_mass_g`? How do you think R summarizes when it encounters missing information?


In R we can get rid of observations with missing information using the function `drop_na()`. We can modify the previous code as:

```{r}
penguins |>
  drop_na(body_mass_g) |>
  group_by(species) |>
  summarize(
    n=n(),
    avg_weight=mean(body_mass_g))

```

Notice that the previous code dropped one observation with missing information from the species Adelie and another from species Gentoo


# Your turn 3: Finding proportions with summarize

1.  When we have categorical variables, we can also calculate proportions!!! Explain why the code below gives the proportion of female penguins among each species.

```{r}
penguins |>
  drop_na(sex) |>
  group_by(species) |>
  summarize(n = n(), prop_female = mean(sex == "female"))
```

2.  Calculate the proportion of each species that live on Biscoe island.

# Your turn 4: Summarizing your data

1. How does the mean `bill_depth_mm` differ across males and females? How about across species?


```{r}
# Solution
penguins |>
  drop_na(sex)|>
  group_by(sex) |>
  summarize(mean_dep = mean(bill_depth_mm))
```


```{r}
penguins |>
  drop_na(bill_depth_mm)|>
  group_by(species) |>
  summarize(mean_dep = mean(bill_depth_mm))

```

2. Do the means of `bill_depth_mm` change a lot across species depending on the year where the data was collected?

```{r}
# Solution
penguins |>
  drop_na(bill_length_mm) |>
  group_by(species, year) |>
  summarize(
    mean_dep = mean(bill_depth_mm),
    sd_dep  = sd(bill_depth_mm)
    )
```


# Your turn 5: More practice!

1.  In the following two chunks, note that all the values in the output are the same, they are just arranged differently. Identify THREE differences in the output. (hint: expand your console window so you can see both outputs at the same time!!!)

```{r}
penguins |>
  group_by(species, sex) |>
  summarize(n = n(), mean_body_mass_g = mean(body_mass_g))
```

```{r}
penguins |>
  group_by(sex, species) |>
  summarize(n = n(), mean_body_mass_g = mean(body_mass_g))
```


2. Predict what the groups will be in the following output BEFORE you run it. Also predict the dimensions of the resulting tibble.

**YOUR GUESS:** Groups=xxx,yyy,zzz. Number of rows=???. Number of columns=???.

```{r}
penguins |>
  group_by(sex, species, island) |>
  summarize(n = n(), mean_body_mass_g = mean(body_mass_g))
```

