---
title: "Tidyverse review"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

# Learning Goals:

- To practice basic transformations from tidyverse

-------------------------------------------------------

```{r, message = FALSE,  warning=FALSE}
library(tidyverse)
```


# Your turn 0:

Let's start our session by remembering what some common functions from the tidyverse do. 

In your groups talk about what the following functions do and briefly summarize them below. After you are done writing, make sure to knit your .Rmd:

* `mutate()`
* `filter()` 
* `summarize()`
* `group_by`


# Analyzing "The Office"

Let's quickly pull our dataset for today, which contains the script information from the iconic show: The Office. 

```{r}
(office_script <- read_csv("~/Sds_101A_SSI26/Class/Data/office_lines.csv"))
```

The first three columns of the dataset correspond to the season, episode number, and episode name for an episode from The Office. The next columns correspond to the `character` who is speaking, the `text` lines that they are speaking, and finally we the `lenght_text` which measures how many letters the `text` has.


# Your turn 1: Practice with the Office

In your groups answer the following questions using R/tidyverse code


1. How many times does Michael speak in the Office during season 1? How about in season 9? Which episode of season 9 featured Michael's comeback and what lines did he say?

2. Start by subsetting your dataset to only lines uttered by Michael, Dwight, Jim, Pam, and Angela, and call this tibble `office_top5`. Use `office_top5` to count how many times lines each character had, and what was their median `length_text`

3. Use `office_top5` to explore the median `length_text` across seasons for characters other than Michael until season 8. Any interesting trends?






